One-Particle Density Matrix Polarisation Susceptibility Tensors

### 2018-08-10 papers/1_dmatpol/submit/paper-submit.tex
Reference and main text check-up final

The theory that is necessary to validate the density matrix susceptibility models from Section~\ref{s:2}
was implemented in our in\hyp{}house plugin to {\sc Psi4} quantum chemistry program.\cite{Psi4.JCTC.2017}
Water molecule was selected as a test system which was subjected to 
electrostatic perturbations due to random uniform electric fields and random collections of point charges
generating non\hyp{}uniform electric fields.
Components of spatially uniform electric fields were determined 
from the uniform distribution in the interval 
of $\vert 0.0404 \vert$ a.u. generating average field of 0.035 a.u. in an ensemble of 100 samples.
Cartesian coordinates of point charges were drawn from the uniform distribution 
of a set of forty three\hyp{}dimensional Euclidean points within a sphere of radius 6.0 Bohrs
with an origin that is located in the centre of mass of water molecule.
In selection of random points, space inside the union of spheres surrounding
oxygen and hydrogen atoms with radii of 4.5 and 4.0 Bohrs, respectively, was
kept free from any point charges due to potential problems with the convergence of the 
self\hyp{}consistent field (SCF) calculations.
Electric charge values were selected from the one\hyp{}dimensional uniform distribution
in the interval of $\vert q_{\rm max} \vert$, where $q_{\rm max}$
was tuned to obtain a desired ensemble of weak, moderate and strong electric field magnitudes
(with $q_{\rm max}=0.05$, $0.2$ and $0.4$ a.u., accordingly).
In this way, independent sets of 100 random samples of either uniform electric field or
point charge distributions were selected and exact solutions of HF\hyp{}SCF equations were 
performed for each validation procedure.
In the case of non\hyp{}uniform electric fields, the distribution centres 
of the \emph{ab initio} DMS model (Eq.~\eqref{e:final-model.HF}) were assumed 
to be LMO centroids, while atoms were used for the generalised DMS models (Eq.~\eqref{e:final-model.General}).
This is because various LMO centroids are often located at the same point which would
introduce linear dependencies in the Hessian matrix from Eq.~\eqref{e:Newton.gH}.

In order to validate all the models discussed in this contribution,
the electric field\hyp{}induced polarisation energy, 
dipole moment and quadrupole moment were analysed in detail.
The electric field\hyp{}induced polarisation energy at the HF level, $\delta E^{\rm Pol}$,
was computed from the following exact formula:
%
\begin{multline}\label{e:dE}
 \delta E^{\rm Pol}[{\bf F}] = 
                     \Tr{\left\{ {\bf D}^{(0)} \cdot
                                \left( 2\delta {\bf J} - \delta {\bf K}  \right) \right\}} \\
                   + \Tr{\left\{ \delta {\bf D} \cdot
                                \left( {\bf H}^{(0)} + {\bf f}^{(0)} + 2{\bf V} + 2\delta {\bf J} - \delta {\bf K}\right) \right\}} \;.
\end{multline}
%
In Eq.~\eqref{e:dE}, ${\bf f}^{(0)}$ and ${\bf H}^{(0)}$ are the Fock matrix and 
the one\hyp{}electron core Hamiltonian matrix of an unperturbed HF ground state,
${\bf V}$ is the one\hyp{}electron external electrostatic potential matrix whereas $\delta {\bf J}$
and $\delta {\bf K}$ are electric field\hyp{}induced changes in the Coulomb and the exchange HF matrices due to 
the polarisation of the electron density, respectively. 
In order to compare the performance of the usual 
distributed dipole\hyp{}dipole polarisability approach,
the approximate interaction energy was also evaluated by using the expression:
%
\begin{equation}\label{e:dE-dpol}
 \delta E^{\rm Pol}[{\bf F}] \approx 
                   - \frac{1}{2} \sum_i^{\rm occ} {\BM \alpha}_i : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_i) \;,
\end{equation}
%
in which ${\BM \alpha}_i$ is the distributed anisotropic dipole\hyp{}dipole polarisability tensor associated with the
$i$th LMO centroid (see Eq.~\eqref{e:dmu-dpol}).
The distributed polarisabilities were computed by utilising the coupled\hyp{}perturbed Hartree\hyp{}Fock
method.\cite{McWeeny.RevModPhys.1960,Dodds.McWeeney.Sadlej.MolPhys.1977} 
Orbital localisation was performed by using the Boys method.\cite{Boys.RevModPhys.1960}

@@

To compute the interaction energy and the induced multipole moments, either exact or approximated forms
of the difference density matrix from Eq.~\eqref{e:dD-Taylor}, Eq.~\eqref{e:final-model.HF} or Eq.~\eqref{e:final-model.General},
were used for a particular electric field distribution. The statistical evaluation of
errors with respect to the benchmark (full HF calculations) were based on 
the following statistical quality descriptors:
%
\begin{enumerate}
 \item Root Mean Square Energy Error (RMSE)
   \begin{equation}\label{e:rmse}
     {\rm RMSE} = \sqrt{ \frac{1}{N_{\rm max}} \sum_N \left( \delta E^{(N)} - \overline{\delta E^{(N)}}\right)^2 } \;;
   \end{equation}
 \item Root Mean Square Dipole Error (RMSD)
   \begin{equation}\label{e:rmsd}
     {\rm RMSD} = \sqrt{ \frac{1}{N_{\rm max}}\sum_N \left( \lvert \delta {\BM\upmu}^{(N)} \rvert 
                                             - \lvert \overline{\delta {\BM\upmu}^{(N)} \rvert }\right)^2 } \;;
   \end{equation}
 \item Root Mean Square Quadrupole Error (RMSQ)
   \begin{equation}\label{e:rmsq}
     {\rm RMSQ} = \sqrt{ \frac{1}{N_{\rm max}}\sum_N \left( \lvert \delta \widetilde{\BM\Theta}^{(N)} \rvert 
                                             - \lvert \overline{\delta \widetilde{\BM\Theta}^{(N)} \rvert }\right)^2 } \;.
   \end{equation}
\end{enumerate}
%
The first describes the ability of the model to reproduce the exact energetics of the system, 
whereas the following two contain additional information on how well the deformation of the one\hyp{}electron
density is described. For Eqs.~\eqref{e:rmsd} and \eqref{e:rmsq} the absolute magnitudes
are defined as
%
\begin{subequations}\label{e:absmagn}
  \begin{align}
   \lvert {\BM\upmu} \rvert &= \sqrt{\mu_x^2+\mu_y^2+\mu_z^2} \;,\\
   \lvert {\widetilde{\BM\Theta}}\rvert &= 
          \sqrt{\widetilde{\Theta}_{zz}^2 + \frac{1}{3}\left(\widetilde{\Theta}_{xx}-\widetilde{\Theta}_{yy}\right)^2 
          + \frac{4}{3}\left( \widetilde{\Theta}_{xy}^2 + \widetilde{\Theta}_{xz}^2 + \widetilde{\Theta}_{yz}^2 \right)} \;. 
  \end{align}
\end{subequations}
%
Throughout the work, the Hartree\hyp{}Fock level\cite{Roothaan.RevModPhys.1951} was assumed 
and the 6-311++G** basis set\cite{Krishnan.Binkley.Seeger.Pople.JCP.1980}
was used. Before all the calculations, water molecule was energy\hyp{}optimised
with a maximum energy, force and displacement thresholds of 10$^{-6}$, 3$\times$$10^{-4}$ 
and 1.2$\times$$10^{-3}$ a.u., respectively.
For all the calculations of the density matrix polarisation susceptibility tensors
from Eqs.~\eqref{e:final-model.HF} and \eqref{e:susceptibility-B}, 
atomic basis was orthogonalised by using the L{\"o}wdin symmetric orthogonalisation scheme,\cite{Mayer.IJQC.2002}
and further back\hyp{}transformed to the non\hyp{}orthogonal AO basis.
To compute numerical first and second derivatives of the one\hyp{}particle density matrix
with respect to the electric field from Eq.~\eqref{e:dD-Taylor}, 
the following finite\hyp{}difference formulae were utilised:
%
\begin{subequations}\label{e:ff}
  \begin{align}
    D_u    &= \frac{D(F_u) - D(-F_u)}{2F_u} \;,\\
    D_{uu} &= \frac{D(F_u) + D(-F_u) - 2D^{(0)}}{F_u^2} \;,\\
    D_{uw} &= \frac{D(F_u,F_w) + D(-F_u,-F_w) + 2D^{(0)}}{2F_uF_w} \nonumber \\ 
           & \; -\frac{D(F_u) + D(-F_u) + D(F_w) + D(-F_w)}{2F_uF_w} \;,
  \end{align}
\end{subequations}
%
where 
the subscripts $u,w=x,y,z$ on the right\hyp{}hand sides denote the derivatives with respect 
to the Cartesian component of the electric field, and the finite difference
steps $F_{u(w)}$ were assumed to be 0.0005~a.u for all components.

@@

Polarisation under the uniform electric field is the simplest type of electrostatic perturbation
whereby it is possible to compute the exact DMS according to Eq.~\eqref{e:dD-Taylor} 
directly by numerical differentiation. Here we consider DMS up to second order, i.e.,
%
\begin{subequations}\label{e:exact-DMS}
 \begin{align}
  {\bf B}^{(10)}_{\alpha\beta} &= \frac{\partial D_{\alpha\beta}}{\partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} \;, \label{e:B10} \\
  {\bf B}^{(20)}_{\alpha\beta} &= \frac{1}{2} 
     \frac{\partial^2 D_{\alpha\beta}}{\partial {\bf F} \otimes \partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} \;, \label{e:B20}
 \end{align}
\end{subequations}
%
and electric field magnitudes in a range from very weak ($<$0.001 a.u.) to very strong ($\sim$0.07 a.u.).
The \emph{ab initio} model from Eq.~\eqref{e:final-model.HF} and
the generalised model from Eq.~\eqref{e:final-model.General} are also discussed for the sake of comparison. 
The associated error distributions in prediction of polarisation energies 
are shown in scatter plot in Figure~\ref{f:fig-1}, whereas the
statistical performance descriptors of polarisation energies and induced multipole moments
are gathered in Table~\ref{t:results}.
%
\begin{figure}[t]
\includegraphics[width=0.5\textwidth]{fig-1.eps}
\caption{\label{f:fig-1} {\bf Performance of the DMS tensors of water molecule
in uniform electric field.} 
$R_{\rm max}$ denotes the maximum rank of the electric field
in the expansion from Eq.~\eqref{e:final-model.General}.
Electric field magnitudes in a range from 0.0 to 0.07 a.u. were tested in 100 samples with randomly selected 
directions of electric field.
} 
\end{figure}
%

@@

%
\begin{table*}%[]
\caption{{\bf Performance of the DMS tensors of water molecule
in uniform and non\hyp{}uniform electric field$^{a,b}$}
}
\label{t:results}
\begin{ruledtabular}
\resizebox{\textwidth}{!}{%
\begin{tabular}{cddddddcddddcdd}
%\hline\hline
\toprule
\multicolumn{1}{l}{} & \multicolumn{6}{c}{\textbf{Ab Initio DMS}} &  & \multicolumn{4}{c}{\textbf{Generalised DMS}}
                                                                  &  & \multicolumn{2}{c}{\textbf{Distributed}} \\
\cline{2-8}
\cline{9-12}
%\cline{14-15}
\textbf{}         
& \multicolumn{2}{c}{$\BM{R_{\rm max}=1}$$^c$}
& \multicolumn{2}{c}{$\BM{R_{\rm max}=1}$$^d$}
& \multicolumn{2}{c}{$\BM{R_{\rm max}=2}$$^d$}
&
& \multicolumn{2}{c}{$\BM{R_{\rm max}=1}$$^e$} 
& \multicolumn{2}{c}{$\BM{R_{\rm max}=2}$$^e$} 
&  
& \multicolumn{2}{c}{\textbf{Polarisability}}        \\ \hline
\multicolumn{15}{c}{} \\ 
\multicolumn{15}{c}{\textbf{Uniform Electric Field}} \\ 
\multicolumn{15}{c}{} \\ \hline
\multicolumn{1}{l}{} & \multicolumn{14}{c}{\textbf{RMSE (kcal/mol)}}              \\ %\hline
   & 4.041 & (n.d) & 4.823 & (n.d.) & 0.231 & (98.40) && 5.0121 & (n.d.) & 0.173 & (99.10) && 0.112 & (99.63) \\ 
\multicolumn{1}{l}{} & \multicolumn{14}{c}{\textbf{RMSD (mD)}}                    \\ %\hline
   & 30.46 & (97.91) & 30.46 & (97.91) & 18.02 & (99.27) && 24.83 & (98.61) & 7.13 & (99.89) && 30.46 & (97.91) \\ 
\multicolumn{1}{l}{} & \multicolumn{14}{c}{\textbf{RMSQ (D\AA $\times 10^{-3}$)}} \\ %\hline
   & 47.89 & (69.17) & 39.24 & (79.30) &  8.56 & (99.02) && 37.63 & (80.96) & 3.94 & (99.79) && 50.66 & (65.49) \\ \hline
\multicolumn{15}{c}{} \\ 
\multicolumn{15}{c}{\textbf{Non-Uniform Electric Field}$^{f}$} \\ 
\multicolumn{15}{c}{} \\ \hline
\multicolumn{1}{l}{} & \multicolumn{14}{c}{\textbf{RMSE (kcal/mol)}}   \\ 
W & 0.067 & (n.d.)  & - &  -  & - &  -  && 0.076 & (n.d.)  & 0.003 & (99.45) && 0.004 & (98.64) \\ 
M & 1.072 & (n.d.)  & - &  -  & - &  -  && 1.236 & (n.d.)  & 0.057 & (99.08) && 0.070 & (98.58) \\ 
S & 4.220 & (n.d.)  & - &  -  & - &  -  && 5.147 & (n.d.)  & 0.402 & (97.25) && 0.332 & (98.13) \\ \hline
\multicolumn{1}{l}{} & \multicolumn{14}{c}{\textbf{RMSD (mD)}}  \\ 
W &  3.36 & (98.71) & - &  -  & - &  -  &&  1.03 & (99.88) &  0.94 & (99.90) &&  3.36 & (98.71) \\ 
M & 14.72 & (98.50) & - &  -  & - &  -  &&  7.72 & (99.59) &  4.55 & (99.86) && 14.72 & (98.50) \\ 
S & 44.80 & (96.92) & - &  -  & - &  -  && 33.06 & (98.33) & 15.96 & (99.61) && 44.80 & (96.92) \\ \hline
\multicolumn{1}{l}{} & \multicolumn{14}{c}{\textbf{RMSQ (D\AA $\times 10^{-3}$)}} \\ 
W &  6.22 & (58.72) & - &  -  & - &  -  &&  1.59 & (97.32) &  1.42 & (97.86) &&  5.45 & (68.29) \\ 
M & 28.22 & (53.90) & - &  -  & - &  -  && 10.45 & (93.68) &  5.68 & (98.13) && 23.94 & (66.83) \\ 
S & 78.07 & (40.53) & - &  -  & - &  -  && 43.33 & (81.68) & 14.40 & (97.98) && 64.33 & (59.62) \\ 
\bottomrule
\end{tabular}%
}
\end{ruledtabular}
%
\footnotesize{\flushleft{
$^a$ $R$-squared coefficients are shown in parentheses. `$n.d.$' stands for `not determined'. 
$^b$ Ranks of the density matrix susceptibility models are denoted as $R_{\rm max}$ 
where $R_{\rm max}$ is the maximum power in the expansion from Eq.~\eqref{e:final-model.General}.
$^c$ \emph{Ab initio} HF model, Eq.~\eqref{e:final-model.HF}.
$^d$ \emph{Ab initio} finite\hyp{}difference model, Eq.~\eqref{e:dD-Taylor}.
$^e$ Linear\hyp{}regression generalised model, Eq.~\eqref{e:final-model.General}.
$^f$ `W', `M' and `S' denote `weak', `moderate' and `strong' electric fields 
that are generated by sets of point charges, respectively (c.f. Table~\ref{t:fields}).\\
}}
\end{table*}

@@

To study the capability of the distributed DMS models to capture 
density matrix polarisation in inhomogeneous electric fields, a water molecule
surrounded by point charges is analysed.
Three different types of point charge environment that generate
weak (0.003 -- 0.007 a.u.), 
moderate (0.01 -- 0.02 a.u.) and strong (0.02 -- 0.05 a.u.) 
electric fields around the water molecule were considered (Table~\ref{t:fields}),
%
\begin{table}[b]
\caption{{\bf Average electric fields in statistical sets of electrostatically perturbed states
of water molecule surrounded by point charges$^a$}
}
\label{t:fields}
\begin{ruledtabular}
\begin{tabular}{ldcdcdcd}
\multirow{2}{*}{\textbf{Set$^b$}} & \multicolumn{7}{c}{\textbf{Average Electric Field (a.u.)}}      \\
                                  & \multicolumn{3}{c}{\textbf{Oxygen Atom}} & \textbf{} 
                                  & \multicolumn{3}{c}{\textbf{Hydrogen Atom}} \\
\cline{2-4}
\cline{6-8}
\textbf{W}                    & 0.0046     & $\pm$     & 0.0019     &           & 0.0046     & $\pm$     & 0.0020     \\
\textbf{M}                    & 0.0184     & $\pm$     & 0.0075     &           & 0.0186     & $\pm$     & 0.0078     \\
\textbf{S}                    & 0.0368     & $\pm$     & 0.0150     &           & 0.0372     & $\pm$     & 0.0156    
\end{tabular}
\end{ruledtabular}
%
\footnotesize{\flushleft{
$^a$ Each set was composed of 100 samples differing in the configuration of 40 charges generated from a uniform distribution.
$^b$ `W', `M' and `S' denote `weak', `moderate' and `strong' electric fields, respectively.\\
}}

>>>>

The theory that is necessary to validate the density matrix susceptibility models from Section~\ref{s:2}
was implemented in our in\hyp{}house plugin to {\sc Psi4} quantum chemistry program.\cite{Psi4.JCTC.2017}
Water molecule was selected as a test system which was subjected to 
electrostatic perturbations due to random uniform electric fields and random collections of point charges
generating non\hyp{}uniform electric fields.
Components of spatially uniform electric fields were determined 
from the uniform distribution in the interval 
of $\vert 0.0404 \vert$ a.u. generating average field of 0.035 a.u. in an ensemble of 100 samples.
Cartesian coordinates of point charges were drawn from the uniform distribution 
of a set of forty three\hyp{}dimensional Euclidean points within a sphere of radius 6.0 Bohrs
with an origin that is located in the centre of mass of water molecule.
In selection of random points, space inside the union of spheres surrounding
oxygen and hydrogen atoms with radii of 4.5 and 4.0 Bohrs, respectively, was
kept free from any point charges due to potential problems with the convergence of the 
self\hyp{}consistent field (SCF) calculations.
Electric charge values were selected from the one\hyp{}dimensional uniform distribution
in the interval of $\vert q_{\rm max} \vert$, where $q_{\rm max}$
was tuned to obtain a desired ensemble of weak, moderate and strong electric field magnitudes
(with $q_{\rm max}=0.05$, $0.2$ and $0.4$ a.u., accordingly).
In this way, independent sets of 100 random samples of either uniform electric field or
point charge distributions were selected and exact solutions of HF\hyp{}SCF equations were 
performed for each validation procedure.
In the case of non\hyp{}uniform electric fields, the distribution centres 
of the \emph{ab initio} DMS model (Eq.~\eqref{e:final-model.HF}) were assumed 
to be the LMO centroids (resulting in the five\hyp{}centre model), 
while the atoms were used for the generalised DMS models (Eq.~\eqref{e:final-model.General}, 
resulting in the three\hyp{}centre models).
This is because certain LMO centroids are often located at the same point which would
introduce linear dependencies in the Hessian matrix from Eq.~\eqref{e:Newton.gH}
when using the generalised DMS scheme.

In order to validate all the models discussed in this contribution,
the electric field\hyp{}induced polarisation energy, 
dipole moment and quadrupole moment were analysed in detail.
The electric field\hyp{}induced polarisation energy at the HF level, $\delta E^{\rm Pol}$,
was computed from the following exact formula:
%
\begin{multline}\label{e:dE}
 \delta E^{\rm Pol}[{\bf F}] = 
                     \Tr{\left\{ {\bf D}^{(0)} \cdot
                                \left( 2\delta {\bf J} - \delta {\bf K}  \right) \right\}} \\
                   + \Tr{\left\{ \delta {\bf D} \cdot
                                \left( {\bf H}^{(0)} + {\bf f}^{(0)} + 2{\bf V} + 2\delta {\bf J} - \delta {\bf K}\right) \right\}} \;.
\end{multline}
%
In Eq.~\eqref{e:dE}, ${\bf f}^{(0)}$ and ${\bf H}^{(0)}$ are the Fock matrix and 
the one\hyp{}electron core Hamiltonian matrix of an unperturbed HF ground state,
${\bf V}$ is the one\hyp{}electron external electrostatic potential matrix whereas $\delta {\bf J}$
and $\delta {\bf K}$ are electric field\hyp{}induced changes in the Coulomb and the exchange HF matrices due to 
the polarisation of the electron density, respectively. 
In order to compare the performance of the usual 
distributed dipole\hyp{}dipole polarisability approach,\cite{Li.Netzloff.Gordon.JCP.2006}
the approximate interaction energy was also evaluated by using the expression:
%
\begin{equation}\label{e:dE-dpol}
 \delta E^{\rm Pol}[{\bf F}] \approx 
                   - \frac{1}{2} \sum_i^{\rm occ} {\BM \alpha}_i : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_i) \;,
\end{equation}
%
in which ${\BM \alpha}_i$ is the distributed anisotropic dipole\hyp{}dipole polarisability tensor associated with the
$i$th LMO centroid (see Eq.~\eqref{e:dmu-dpol}).
The distributed polarisabilities were computed by utilising the coupled\hyp{}perturbed Hartree\hyp{}Fock
method.\cite{McWeeny.RevModPhys.1960,Dodds.McWeeney.Sadlej.MolPhys.1977} 
Orbital localisation was performed by using the Boys method.\cite{Boys.RevModPhys.1960}

@@

To compute the interaction energy and the induced multipole moments, either exact or approximated forms
of the difference density matrix from Eq.~\eqref{e:dD-Taylor}, Eq.~\eqref{e:final-model.HF} or Eq.~\eqref{e:final-model.General},
were used for a particular electric field distribution. The statistical evaluation of
errors with respect to the benchmark (full HF calculations) were based on 
the following statistical quality descriptors:
%
\begin{enumerate}
 \item Root Mean Square Energy Error (RMSE)
   \begin{equation}\label{e:rmse}
     {\rm RMSE} = \sqrt{ \frac{1}{N_{\rm max}} \sum_N^{N_{\rm max}} \left( \delta E^{(N)} - \overline{\delta E^{(N)}}\right)^2 } \;;
   \end{equation}
 \item Root Mean Square Dipole Error (RMSD)
   \begin{equation}\label{e:rmsd}
     {\rm RMSD} = \sqrt{ \frac{1}{N_{\rm max}}\sum_N^{N_{\rm max}} \left( \lvert \delta {\BM\upmu}^{(N)} \rvert 
                                             - \lvert \overline{\delta {\BM\upmu}^{(N)} \rvert }\right)^2 } \;;
   \end{equation}
 \item Root Mean Square Quadrupole Error (RMSQ)
   \begin{equation}\label{e:rmsq}
     {\rm RMSQ} = \sqrt{ \frac{1}{N_{\rm max}}\sum_N^{N_{\rm max}} \left( \lvert \delta \widetilde{\BM\Theta}^{(N)} \rvert 
                                             - \lvert \overline{\delta \widetilde{\BM\Theta}^{(N)} \rvert }\right)^2 } \;.
   \end{equation}
\end{enumerate}
%
The first describes the ability of the model to reproduce the exact energetics of the system, 
whereas the following two contain additional information on how well the deformation of the one\hyp{}electron
density is described. For Eqs.~\eqref{e:rmsd} and \eqref{e:rmsq} the absolute magnitudes
are defined as
%
\begin{subequations}\label{e:absmagn}
  \begin{align}
   \lvert {\BM\upmu} \rvert &= \sqrt{\mu_x^2+\mu_y^2+\mu_z^2} \;,\\
   \lvert {\widetilde{\BM\Theta}}\rvert &= 
          \sqrt{\widetilde{\Theta}_{zz}^2 + \frac{1}{3}\left(\widetilde{\Theta}_{xx}-\widetilde{\Theta}_{yy}\right)^2 
          + \frac{4}{3}\left( \widetilde{\Theta}_{xy}^2 + \widetilde{\Theta}_{xz}^2 + \widetilde{\Theta}_{yz}^2 \right)} \;. 
  \end{align}
\end{subequations}
%
Throughout the work, the Hartree\hyp{}Fock level\cite{Roothaan.RevModPhys.1951} was assumed 
and the 6-311++G** basis set\cite{Krishnan.Binkley.Seeger.Pople.JCP.1980}
was used. Before all the calculations, water molecule was energy\hyp{}optimised
with a maximum energy, force and displacement thresholds of 10$^{-6}$, 3$\times$$10^{-4}$ 
and 1.2$\times$$10^{-3}$ a.u., respectively.
For all the calculations of the density matrix polarisation susceptibility tensors
from Eqs.~\eqref{e:final-model.HF} and \eqref{e:susceptibility-B}, 
atomic basis was orthogonalised by using the L{\"o}wdin symmetric orthogonalisation scheme,\cite{Mayer.IJQC.2002}
and further back\hyp{}transformed to the non\hyp{}orthogonal AO basis.
To compute numerical first and second derivatives of the one\hyp{}particle density matrix
with respect to the electric field from Eq.~\eqref{e:dD-Taylor}, 
the following finite\hyp{}difference formulae were utilised:
%
\begin{subequations}\label{e:ff}
  \begin{align}
    D_u    &= \frac{D(F_u) - D(-F_u)}{2F_u} \;,\\
    D_{uu} &= \frac{D(F_u) + D(-F_u) - 2D^{(0)}}{F_u^2} \;,\\
    D_{uw} &= \frac{D(F_u,F_w) + D(-F_u,-F_w) + 2D^{(0)}}{2F_uF_w} \nonumber \\ 
           & \; -\frac{D(F_u) + D(-F_u) + D(F_w) + D(-F_w)}{2F_uF_w} \;,
  \end{align}
\end{subequations}
%
where 
the subscripts $u,w=x,y,z$ on the right\hyp{}hand sides denote the derivatives with respect 
to the Cartesian component of the electric field, and the finite difference
steps $F_{u(w)}$ were assumed to be 0.0005~a.u. for all components.

@@

Polarisation under the uniform electric field is the simplest type of electrostatic perturbation
whereby it is possible to compute the exact DMS according to Eq.~\eqref{e:dD-Taylor} 
directly by numerical differentiation. Here we consider DMS up to second order, i.e.,
%
\begin{subequations}\label{e:exact-DMS}
 \begin{align}
  {\bf B}^{(10)}_{\alpha\beta} &= \frac{\partial D_{\alpha\beta}}{\partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} \;, \label{e:B10} \\
  {\bf B}^{(20)}_{\alpha\beta} &= \frac{1}{2} 
     \frac{\partial^2 D_{\alpha\beta}}{\partial {\bf F} \otimes \partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} \;, \label{e:B20}
 \end{align}
\end{subequations}
%
and electric field magnitudes in a range from very weak ($<$0.001 a.u.) to very strong ($\sim$0.07 a.u.).
The \emph{ab initio} model from Eq.~\eqref{e:final-model.HF} and
the generalised model from Eq.~\eqref{e:final-model.General} are also discussed for the sake of comparison. 
The associated error distributions in prediction of polarisation energies 
are shown in scatter plot in Figure~\ref{f:fig-1}, whereas the
statistical performance descriptors of polarisation energies and induced multipole moments
are gathered in Table~\ref{t:results}.
%
\begin{figure}[t]
\includegraphics[width=0.5\textwidth]{fig-1.eps}
\caption{\label{f:fig-1} {\bf Performance of the DMS tensors of the ground\hyp{}state water molecule
in uniform electric field.} 
$R_{\rm max}$ denotes the maximum rank of the electric field
in the expansion from Eq.~\eqref{e:final-model.General}.
Electric field magnitudes in a range from 0.0 to 0.07 a.u. were tested in 100 samples with randomly selected 
directions of electric field.
} 
\end{figure}
%

@@

%
\begin{table*}%[]
\caption{{\bf Performance of the DMS tensors of ground\hyp{}state water molecule
in uniform and non\hyp{}uniform electric field$^{a,b}$}
}
\label{t:results}
\begin{ruledtabular}
\resizebox{\textwidth}{!}{%
\begin{tabular}{cddddddcddddcdd}
%\hline\hline
\toprule
\multicolumn{1}{l}{} & \multicolumn{6}{c}{\textbf{Ab Initio DMS}} &  & \multicolumn{4}{c}{\textbf{Generalised DMS}}
                                                                  &  & \multicolumn{2}{c}{\textbf{Distributed}} \\
\cline{2-8}
\cline{9-12}
%\cline{14-15}
\textbf{}         
& \multicolumn{2}{c}{$\BM{R_{\rm max}=1}$$^c$}
& \multicolumn{2}{c}{$\BM{R_{\rm max}=1}$$^d$}
& \multicolumn{2}{c}{$\BM{R_{\rm max}=2}$$^d$}
&
& \multicolumn{2}{c}{$\BM{R_{\rm max}=1}$$^e$} 
& \multicolumn{2}{c}{$\BM{R_{\rm max}=2}$$^e$} 
&  
& \multicolumn{2}{c}{\textbf{Polarisability}}        \\ \hline
\multicolumn{15}{c}{} \\ 
\multicolumn{15}{c}{\textbf{Uniform Electric Field}} \\ 
\multicolumn{15}{c}{} \\ \hline
\multicolumn{1}{l}{} & \multicolumn{14}{c}{\textbf{RMSE (kcal/mol)}}              \\ %\hline
   & 4.041 & (n.d) & 4.823 & (n.d.) & 0.231 & (98.40) && 5.0121 & (n.d.) & 0.173 & (99.10) && 0.112 & (99.63) \\ 
\multicolumn{1}{l}{} & \multicolumn{14}{c}{\textbf{RMSD (mD)}}                    \\ %\hline
   & 30.46 & (97.91) & 30.46 & (97.91) & 18.02 & (99.27) && 24.83 & (98.61) & 7.13 & (99.89) && 30.46 & (97.91) \\ 
\multicolumn{1}{l}{} & \multicolumn{14}{c}{\textbf{RMSQ (D\AA $\times 10^{-3}$)}} \\ %\hline
   & 47.89 & (69.17) & 39.24 & (79.30) &  8.56 & (99.02) && 37.63 & (80.96) & 3.94 & (99.79) && 50.66 & (65.49) \\ \hline
\multicolumn{15}{c}{} \\ 
\multicolumn{15}{c}{\textbf{Non-Uniform Electric Field}$^{f}$} \\ 
\multicolumn{15}{c}{} \\ \hline
\multicolumn{1}{l}{} & \multicolumn{14}{c}{\textbf{RMSE (kcal/mol)}}   \\ 
W & 0.067 & (n.d.)  & - &  -  & - &  -  && 0.076 & (n.d.)  & 0.003 & (99.45) && 0.004 & (98.64) \\ 
M & 1.072 & (n.d.)  & - &  -  & - &  -  && 1.236 & (n.d.)  & 0.057 & (99.08) && 0.070 & (98.58) \\ 
S & 4.220 & (n.d.)  & - &  -  & - &  -  && 5.147 & (n.d.)  & 0.402 & (97.25) && 0.332 & (98.13) \\ \hline
\multicolumn{1}{l}{} & \multicolumn{14}{c}{\textbf{RMSD (mD)}}  \\ 
W &  3.36 & (98.71) & - &  -  & - &  -  &&  1.03 & (99.88) &  0.94 & (99.90) &&  3.36 & (98.71) \\ 
M & 14.72 & (98.50) & - &  -  & - &  -  &&  7.72 & (99.59) &  4.55 & (99.86) && 14.72 & (98.50) \\ 
S & 44.80 & (96.92) & - &  -  & - &  -  && 33.06 & (98.33) & 15.96 & (99.61) && 44.80 & (96.92) \\ \hline
\multicolumn{1}{l}{} & \multicolumn{14}{c}{\textbf{RMSQ (D\AA $\times 10^{-3}$)}} \\ 
W &  6.22 & (58.72) & - &  -  & - &  -  &&  1.59 & (97.32) &  1.42 & (97.86) &&  5.45 & (68.29) \\ 
M & 28.22 & (53.90) & - &  -  & - &  -  && 10.45 & (93.68) &  5.68 & (98.13) && 23.94 & (66.83) \\ 
S & 78.07 & (40.53) & - &  -  & - &  -  && 43.33 & (81.68) & 14.40 & (97.98) && 64.33 & (59.62) \\ 
\bottomrule
\end{tabular}%
}
\end{ruledtabular}
%
\footnotesize{\flushleft{
$^a$ $R$-squared coefficients are shown in parentheses. `$n.d.$' stands for `not determined'. 
$^b$ Ranks of the density matrix susceptibility models are denoted as $R_{\rm max}$ 
where $R_{\rm max}$ is the maximum power in the expansion from Eq.~\eqref{e:final-model.General}.
$^c$ \emph{Ab initio} HF model, Eq.~\eqref{e:final-model.HF}.
$^d$ \emph{Ab initio} finite\hyp{}difference model, Eq.~\eqref{e:dD-Taylor}.
$^e$ Linear\hyp{}regression generalised model, Eq.~\eqref{e:final-model.General}.
$^f$ `W', `M' and `S' denote `weak', `moderate' and `strong' electric fields 
that are generated by sets of point charges, respectively (c.f. Table~\ref{t:fields}).\\
}}
\end{table*}

@@

To study the capability of the distributed DMS models to capture 
density matrix polarisation in inhomogeneous electric fields, a water molecule
surrounded by point charges is analysed.
Three different types of point charge environment that generate
weak (0.003 -- 0.007 a.u.), 
moderate (0.01 -- 0.02 a.u.) and strong (0.02 -- 0.05 a.u.) 
electric fields around the water molecule were considered (Table~\ref{t:fields}),
%
\begin{table}[b]
\caption{{\bf Average electric fields in statistical sets of electrostatically perturbed ground states
of water molecule surrounded by point charges$^a$}
}
\label{t:fields}
\begin{ruledtabular}
\begin{tabular}{ldcdcdcd}
\multirow{2}{*}{\textbf{Set$^b$}} & \multicolumn{7}{c}{\textbf{Average Electric Field (a.u.)}}      \\
                                  & \multicolumn{3}{c}{\textbf{Oxygen Atom}} & \textbf{} 
                                  & \multicolumn{3}{c}{\textbf{Hydrogen Atom}} \\
\cline{2-4}
\cline{6-8}
\textbf{W}                    & 0.0046     & $\pm$     & 0.0019     &           & 0.0046     & $\pm$     & 0.0020     \\
\textbf{M}                    & 0.0184     & $\pm$     & 0.0075     &           & 0.0186     & $\pm$     & 0.0078     \\
\textbf{S}                    & 0.0368     & $\pm$     & 0.0150     &           & 0.0372     & $\pm$     & 0.0156    
\end{tabular}
\end{ruledtabular}
%
\footnotesize{\flushleft{
$^a$ Each set was composed of 100 samples differing in the configuration of 40 charges generated from a uniform distribution.
$^b$ `W', `M' and `S' denote `weak', `moderate' and `strong' electric fields, respectively.\\
}}

\end{table}
%
because similar ranges of electric field are experienced by water molecules
in liquid phase.\cite{Fried.Wang.Boxer.Ren.Pande.JPCB.2013,Reischl.Kofinger.Dellago.MolPhys.2009}
Performance descriptors of the \emph{ab initio} and generalised 
models of the density matrix susceptibility, as well as 
the conventional dipole\hyp{}dipole distributed polarisability
approach, are shown in Table~\ref{t:results}, whereas the error distributions
in strong electric fields are additionally plotted in Figure~\ref{f:fig-2}A--I.
